Neural Circuit Organization of the Ctenophore Aboral Organ

### 2025-03-24 manuscript/Manuscript.qmd
MS: main text

---
title: "Neural Circuit Organization of the Ctenophore Apical Organ"

@@

## Abstract
<!--# I rewrote a bit the abstract, please check -->
Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth. During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns. To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the apical organ of the ctenophore *Mnemiopsis leidyi*. We reconstructed 909 cells, including syncytial neurons that form a nerve net and synapse on the balancer cells. Our connectome analysis revealed that these syncytial neurons establish direct and indirect pathways to balancer cells, with "bridge cells" acting as relay neurons. High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation. Given that ctenophores branched off early from other animals, these results inform our understanding of the early divergence of nervous systems.

@@

Ctenophores show sophisticated and dynamic behavioral patterns and respond to a wide range of external stimuli. These behaviors are thought to be regulated by multiple neural systems, including the subepithelial nerve net (SNN), mesogleal neurons, an apical organ, and tentacular nerves. However, the mechanisms by which these neural systems control behavior remain largely unexplored.
Molecular phylogeny and chromosome-level synteny analyses suggest that ctenophores may be the sister group to all other animals [@Ryan2013; @Whelan2015a; @Whelan2017; @Whelan2023; @Li2021; @Schultz2023]. In agreement with this early divergence from all other animals with a nervous system, the ctenophore nervous system shows several unique features that suggest independent evolution from other animals <!--# cite some Moroz papers -->.

@@

Another unique feature of ctenophores is their aboral organ that contains the statocyst. Ultrastructural analyses have revealed that the aboral organ contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974; @Tamm1982]. Immunostaining with an antibody labelling tyrosylated α-tubulin identified a neural structure called the deep nerve net within the apical organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].

@@

The apical organ, also known as aboral organ, of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014]. The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014; @Tamm1982]. Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron].

@@

In this study, we combine volume EM and high-speed imaging to examine the gravity-sensing system within the apical organ of the ctenophore *Mnemiopsis leidyi*. Through connectome analysis, we uncovered the structure and synaptic organisation of the aboral nerve net. Our imaging experiments revealed correlated changes in the activity of balancer cilia, suggesting a coordinating activity for the aboral nerve net.

@@

### Volume EM reconstruction of the *Mnemiopsis* apical organ

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the apical organ. Subsequently, 620 of these images were stitched and aligned in TrakEM2 <!--# add reference-->. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated  all cells in this dataset in CATMAID <!--#  ref-->, resulting in a reconstruction of 909 cells with nuclei and cell bodies intact. Most of the cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.

@@

Next, we divided the entire apical organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F)<!--#  add reference to figures also elsewhere-->. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores <!--#  ref-->. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development<!--# ref -->. We assigned each of the 909 skeletonised cells to one of the four quadrants.

![**Figure 1. Morphology of the apical organ in *Mnemiopsis leidyi***\
(A) Whole-body image of a 5-day-old *M. leidyi* cydippid larva in lateral view. The boxed region indicates the apical organ (ao). Scale bar: 100 µm.\
(B) Schematic diagram of the aboral view of a cydippid larva. The two-tone coloration represents the biradial symmetry of the body. For convenience, the body is divided into four quadrants, designated as the first (Q1) through fourth (Q4) quadrants. The two primary viewing angles are referred to as the sagittal plane (S) and the tentacular plane (T). Abbreviations: ao, apical organ; cg, ciliated groove; cr, comb plates; tb, tentacle bulbs.\
(C) Enlarged views of the apical organ from different perspectives in a cydippid larva using a DIC microscope. The left panel presents an aboral view of the apical organ. The middle panel shows a lateral view of the apical organ in the sagittal (S) plane, while the right panel displays a lateral view in the tentacular (T) plane. Abbreviations: bal, balancer; cg, ciliated grooves; do, dome cilia; li, lithocyte. Scale bar: 50 µm.\
(D) Schematic representations of the apical organ from three different perspectives. The left panel illustrates the aboral view, the middle panel shows the lateral view in the sagittal plane, and the right panel presents the lateral view in the tentacular plane. The quadrant colors correspond to those in panel (B).\
(E) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets. The spherical objects indicate nuclear positions, while the lines represent the traced cell centers. Scale bar: 10 µm.\
(F) Three-dimensional reconstruction of cells composing the apical organ, displayed from different perspectives. The left panel presents an aboral view of the apical organ. The middle panel shows a sagittal plane view, while the right panel provides a tentacular plane view. Cells are color-coded according to their respective quadrants. Lithocytes (li) are represented as three gray spheres, while balancers (bal) are depicted as gray lines. Scale bar: 25 µm.](figures/Figure1.png)

@@

Classical neural staining techniques did not provide clear images of neurons at the aboral pole. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor <!--# what is this exactly? are there epithelial cells? -->of the apical organ [@Horridge_1964; @Hernandez_Nicaise_1973]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) <!--# figure.. -->. We could not detect specialized post-synaptic structures, in agreement with previous studies <!--# refs -->. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells <!--# figure of the distribution -->.

@@

![**Figure 2. Organisation of the aboral synaptic nerve net**\
(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (blue). The spheres represent the positions of individual nuclei. The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei. The spheres indicate the positions of the nuclei. The left panel shows a aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Graph showing the number of mitochondria per cell and the proportion of mitochondria involved in synapse formation for each cell type. Green represents mitochondria associated with synapses, while gray represents mitochondria not involved in synapse formation. ANN has a significantly higher number of mitochondria compared to other cell types, with most of them forming synapses. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(D) Localization of mitochondria within ANN Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). The abbreviations are as follows: mi (mitochondrion), er (endoplasmic reticulum), and sv (synaptic vesicles).\
(F) Plot showing the three-dimensional positions of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN Q1-4.\
(G) Neural circuit map of ANN connectivity. The blue, orange, and pink circles represent individual neurons in ANN Q1-4, Q2Q3, and Q3Q4, respectively. Magenta arrows indicate synaptic connections from the center ANN to the lateral ANN, while cyan arrows indicate connections from the lateral ANN to the center ANN. Numbers denote the number of synapses. Blue arrows represent autapses in ANN Q1-4 along with their synapse counts.](figures/Figure2.png)

@@

Balancer cells are monociliated cells with long motile cilia that form four bundles of compound cilia, one in each quadrant. These bundles come together at the centre of the apical organ to support the cellular mass of the statolyth. The bending of the cilia and the position of somata differed clearly when viewed laterally from the sagittal or the tentacular plane (Figure 3B). In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28). Each cell contained 3 to 10 <!-- check --> mitochondria.

@@

![**Figure 3. Reconstruction of the balancers and bridges that receives synapses from the ANN.**\
(A) Graph showing the number of synaptic outputs from the ANN circuit for each cell type. The balancer and bridge cell types receive relatively high outputs from the ANN circuit. ANN represents the number of synapses within the ANN circuit. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer ciliated cells (light blue). The thickness of the arrows and the numbers correspond to the number of synapses. Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.\
(C) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The morphology of individual bridge cells extending across opposite quadrant regions is depicted. The spheres represent the positions of individual nuclei. The left panel shows an aboral view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(D) Mitochondrial localization within bridge cells and associated presynaptic triad structures. Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified. The left panel shows a dorsal view of the apical organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(E) Synaptic connections from ANN neurons to balancer ciliated cells. The positions of synapses from ANNs to balancers (magenta) are indicated. The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view. Balancer ciliated cells are depicted in light gray.\
(F) Synaptic connections between ANNs and bridge cells. The positions of synapses from ANNs to bridge cells (magenta) and from bridge cells to ANNs (light blue) are indicated. The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view. Bridge cells are shown in light gray.\
(G) Connectivity matrix of the gravity-sensing neural circuit. Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups. The numbers and varying shades of blue correspond to the number of synapses.\
(H) Complete connectivity map of the gravity-sensing neural circuit. Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels. The thickness of the arrows and the numerical values indicate the number of synapses. ANN Q1-4 are shown in blue, ANN Q1Q2 in orange, ANNs in Q3Q4 in pink, balancer cell groups in gray, and bridge cell groups in yellow.](figures/Figure3.png)

@@

We identified three syncytial neurons with multiple nuclei in the apical organ of five-day-old *M. leidyi* cydippid larvae. Burkhurdt et al. previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia [@Burkhardt2023]. However, the newly discovered syncytial neurons embedded within the apical organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons. Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa [@Mayorova2021], with cell bodies extending through intercellular spaces. These morphological characteristics suggest that the newly identified neurons represent a second distinct type of syncytial neuron in ctenophores.

@@

In the this study, we found that three neurons are closely positioned within the confined space of the apical organ, potentially enabling a single neuron to control distant ciliary movements in a synchronized manner. Moreover, the presence of neurons projecting to the same ciliated cells in distinct patterns suggests that these neurons may serve opposing functions—one inhibiting ciliary activity while the other promotes it, or one modulating ciliary beat frequency. While our synaptic ultrastructure analysis did not allow us to distinguish excitatory from inhibitory neurons, future electrophysiological investigations will be essential to elucidate their functional properties.

@@

Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e; @Krisch1973; Aronova1974]. By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the apical organ. However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets. In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the apical organ [@Sachkova2021; @Hayakawa2022]. These findings suggest that many sensory cells in the apical organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis [@Jékely2021]. While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.

@@

Surprisingly, our study identified bridge cells as the primary cell type forming synaptic inputs onto neurons in the apical organ. Bridge cells were first described by Tamm et al. based on their morphological characteristics, with cellular projections extending to the basal region of balancer cilia [@Tamm_2002]. Our findings suggest that bridge cells are electrically coupled to balancer cilia and provide rapid feedback to neurons via synaptic transmission. Future studies analyzing differential responses to various sensory stimuli will not only elucidate the functional roles of bridge cells but also provide insights into the evolutionary relationship between transmission speed and synapse development.

@@

The neural cells identified in the apical organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets. Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution [@Luo2021]. Moreover, the presence of multiple divergent feedforward circuits suggests that slight variations in the synchronization and phase delay of ciliary beating (within the range of a few hundred milliseconds) may have evolved as a mechanism for generating behavioral diversity.

>>>>

---
title: "Neural Circuit Organization of the Ctenophore Aboral Organ"

@@

## Abstract

<!--# I rewrote a bit the abstract, please check -->

Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth. During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns. To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*. We reconstructed 909 cells, including syncytial neurons that form a nerve net and synapse on the balancer cells. Our connectome analysis revealed that these syncytial neurons establish direct and indirect pathways to balancer cells, with "bridge cells" acting as relay neurons. High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation. Given that ctenophores branched off early from other animals, these results inform our understanding of the early divergence of nervous systems.

@@

Ctenophores show sophisticated and dynamic behavioral patterns and respond to a wide range of external stimuli. These behaviors are thought to be regulated by multiple neural systems, including the subepithelial nerve net (SNN), mesogleal neurons, an aboral organ, and tentacular nerves. However, the mechanisms by which these neural systems control behavior remain largely unexplored. Molecular phylogeny and chromosome-level synteny analyses suggest that ctenophores may be the sister group to all other animals [@Ryan2013; @Whelan2015a; @Whelan2017; @Whelan2023; @Li2021; @Schultz2023]. In agreement with this early divergence from all other animals with a nervous system, the ctenophore nervous system shows several unique features that suggest independent evolution from other animals <!--# cite some Moroz papers -->.

@@

Another unique feature of ctenophores is their aboral organ that contains the statocyst. Ultrastructural analyses have revealed that the aboral organ contains a diverse array of sensory cells and neuronal components [@Hernandez-Nicaise1968; @Hernandez_Nicaise_1973; @Hernandez-Nicaise1974; @Tamm1982]. Immunostaining with an antibody labelling tyrosylated α-tubulin identified a neural structure called the deep nerve net within the aboral organ, suggesting its involvement in neural regulation of sensory-motor control [@Jager2011].

@@

The aboral organ, also known as aboral organ, of ctenophores contains a dense mineralized mass known as the statocyst, supported by four mechanosensory ciliated cells called balancer cells [@Noda_2014]. The ciliary movements of these balancer cells function as pacemakers for the eight rows of comb plates that control posture and swimming through mechanotransduction [@Tamm_2014; @Tamm1982]. Synaptic connections projecting to the balancer cells have also been reported [@aronova1974electron].

@@

In this study, we combine volume EM and high-speed imaging to examine the gravity-sensing system within the aboral organ of the ctenophore *Mnemiopsis leidyi*. Through connectome analysis, we uncovered the structure and synaptic organisation of the aboral nerve net. Our imaging experiments revealed correlated changes in the activity of balancer cilia, suggesting a coordinating activity for the aboral nerve net.

@@

### Volume EM reconstruction of the *Mnemiopsis* aboral organ

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, 620 of these images were stitched and aligned in TrakEM2 <!--# add reference-->. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID <!--#  ref-->, resulting in a reconstruction of 909 cells with nuclei and cell bodies intact. Most of the cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ.

@@

Next, we divided the entire aboral organ into four broad quadrants (Q1-Q4) to facilitate grouping the identified cells (Figure 1F)<!--#  add reference to figures also elsewhere-->. The general body plan of ctenophores, when viewed from the aboral side, exhibits biradial symmetry around the anal pores <!--#  ref-->. This symmetry corresponds to the four blastomeres present at the four-cell stage during early embryonic development<!--# ref -->. We assigned each of the 909 skeletonised cells to one of the four quadrants.

![**Figure 1. Morphology of the aboral organ in *Mnemiopsis leidyi***\
(A) Whole-body image of a 5-day-old *M. leidyi* cydippid larva in lateral view. The boxed region indicates the aboral organ (ao). Scale bar: 100 µm.\
(B) Schematic diagram of the aboral view of a cydippid larva. The two-tone coloration represents the biradial symmetry of the body. For convenience, the body is divided into four quadrants, designated as the first (Q1) through fourth (Q4) quadrants. The two primary viewing angles are referred to as the sagittal plane (S) and the tentacular plane (T). Abbreviations: ao, aboral organ; cg, ciliated groove; cr, comb plates; tb, tentacle bulbs.\
(C) Enlarged views of the aboral organ from different perspectives in a cydippid larva using a DIC microscope. The left panel presents an aboral view of the aboral organ. The middle panel shows a lateral view of the aboral organ in the sagittal (S) plane, while the right panel displays a lateral view in the tentacular (T) plane. Abbreviations: bal, balancer; cg, ciliated grooves; do, dome cilia; li, lithocyte. Scale bar: 50 µm.\
(D) Schematic representations of the aboral organ from three different perspectives. The left panel illustrates the aboral view, the middle panel shows the lateral view in the sagittal plane, and the right panel presents the lateral view in the tentacular plane. The quadrant colors correspond to those in panel (B).\
(E) Example of cell tracing using the collaborative annotation toolkit CATMAID for large-scale electron microscopy image datasets. The spherical objects indicate nuclear positions, while the lines represent the traced cell centers. Scale bar: 10 µm.\
(F) Three-dimensional reconstruction of cells composing the aboral organ, displayed from different perspectives. The left panel presents an aboral view of the aboral organ. The middle panel shows a sagittal plane view, while the right panel provides a tentacular plane view. Cells are color-coded according to their respective quadrants. Lithocytes (li) are represented as three gray spheres, while balancers (bal) are depicted as gray lines. Scale bar: 25 µm.](figures/Figure1.png)

@@

Classical neural staining techniques did not provide clear images of neurons at the aboral pole. However, ultrastructural studies revealed morphological evidence for neural elements, including synapses that are located on the epithelial floor <!--# what is this exactly? are there epithelial cells? -->of the aboral organ [@Horridge_1964; @Hernandez_Nicaise_1973]. In our dataset, we identified synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) <!--# figure.. -->. We could not detect specialized post-synaptic structures, in agreement with previous studies <!--# refs -->. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells <!--# figure of the distribution -->.

@@

![**Figure 2. Organisation of the aboral synaptic nerve net**\
(A) 3D reconstruction of the aboral nerve net (ANN) Q1Q2Q3Q4 (Q1-4), which spans all four quadrants and contains multiple nuclei (blue). The spheres represent the positions of individual nuclei. The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(B) 3D reconstruction of two ANN Q1Q2 (pink) and Q3Q4 (orange), each spanning two quadrants and containing multiple nuclei. The spheres indicate the positions of the nuclei. The left panel shows a aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(C) Graph showing the number of mitochondria per cell and the proportion of mitochondria involved in synapse formation for each cell type. Green represents mitochondria associated with synapses, while gray represents mitochondria not involved in synapse formation. ANN has a significantly higher number of mitochondria compared to other cell types, with most of them forming synapses. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(D) Localization of mitochondria within ANN Q1-4. Red indicates mitochondria associated with the presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified.\
(E) Representative electron micrographs of presynaptic triad structures observed in the dataset and their schematic diagrams. The left diagram illustrates synaptic projections from the center ANN (ANN Q1-4) to the lateral ANN (ANN Q3Q4), while the right diagram shows synaptic projections from the lateral ANN (ANN Q1Q2) to the center ANN (ANN Q1-4). The abbreviations are as follows: mi (mitochondrion), er (endoplasmic reticulum), and sv (synaptic vesicles).\
(F) Plot showing the three-dimensional positions of synapses. Synapses from the center ANN to the lateral ANN are shown in magenta, while synapses from the lateral ANN to the center ANN are shown in cyan. Blue dots indicate the locations of autapses within ANN Q1-4.\
(G) Neural circuit map of ANN connectivity. The blue, orange, and pink circles represent individual neurons in ANN Q1-4, Q2Q3, and Q3Q4, respectively. Magenta arrows indicate synaptic connections from the center ANN to the lateral ANN, while cyan arrows indicate connections from the lateral ANN to the center ANN. Numbers denote the number of synapses. Blue arrows represent autapses in ANN Q1-4 along with their synapse counts.](figures/Figure2.png)

@@

Balancer cells are monociliated cells with long motile cilia that form four bundles of compound cilia, one in each quadrant. These bundles come together at the centre of the aboral organ to support the cellular mass of the statolyth. The bending of the cilia and the position of somata differed clearly when viewed laterally from the sagittal or the tentacular plane (Figure 3B). In each quadrant, there were between 28-37 balancer cells (Q1: 37; Q2: 32; Q3: 32; Q4: 28). Each cell contained 3 to 10 <!-- check --> mitochondria.

@@

![**Figure 3. Reconstruction of the balancers and bridges that receives synapses from the ANN.**\
(A) Graph showing the number of synaptic outputs from the ANN circuit for each cell type. The balancer and bridge cell types receive relatively high outputs from the ANN circuit. ANN represents the number of synapses within the ANN circuit. The abbreviations are as follows: bal (balancer), brg (bridge), bsl (bristle), cg (ciliated groove), dv (dense vesicle cells), imc (intra-multiciliated cells), la (lamellate bodies), li (lithocytes), pl (plumose), and ef (epithelial floor cells).\
(B) Connectivity map of three ANN neurons (Q1-4 in blue, Q1Q2 in pink, and Q3Q4 in orange) and balancer ciliated cells (light blue). The thickness of the arrows and the numbers correspond to the number of synapses. Synaptic structures from the same neuron targeting the same balancer cell are grouped into hexagons, with the number of cells added to the label.\
(C) 3D reconstruction of bridge cells spanning the Q1Q2 and Q3Q4 quadrants. The Q1Q2-side bridge cells (8 cells) are shown in blue, while the Q3Q4-side bridge cells (6 cells) are shown in xxx color. The morphology of individual bridge cells extending across opposite quadrant regions is depicted. The spheres represent the positions of individual nuclei. The left panel shows an aboral view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(D) Mitochondrial localization within bridge cells and associated presynaptic triad structures. Red indicates mitochondria associated with presynaptic triad structures, yellow marks mitochondria containing synaptic vesicles but lacking a clearly defined presynaptic triad, blue represents mitochondria with unclear synaptic vesicles, and black denotes mitochondria where no synaptic vesicles were identified. The left panel shows a dorsal view of the aboral organ, the middle panel presents a sagittal plane view, and the right panel provides a tentacular plane view.\
(E) Synaptic connections from ANN neurons to balancer ciliated cells. The positions of synapses from ANNs to balancers (magenta) are indicated. The left panel presents an aboral view, while the right panel shows a lateral view and a tentacular plane view. Balancer ciliated cells are depicted in light gray.\
(F) Synaptic connections between ANNs and bridge cells. The positions of synapses from ANNs to bridge cells (magenta) and from bridge cells to ANNs (light blue) are indicated. The left panel shows an aboral view, while the right panel presents a tentacular plane lateral view. Bridge cells are shown in light gray.\
(G) Connectivity matrix of the gravity-sensing neural circuit. Columns represent presynaptic cell groups, while rows represent postsynaptic cell groups. The numbers and varying shades of blue correspond to the number of synapses.\
(H) Complete connectivity map of the gravity-sensing neural circuit. Cells belonging to the same group are enclosed in hexagons, and the number of cells is added to their labels. The thickness of the arrows and the numerical values indicate the number of synapses. ANN Q1-4 are shown in blue, ANN Q1Q2 in orange, ANNs in Q3Q4 in pink, balancer cell groups in gray, and bridge cell groups in yellow.](figures/Figure3.png)

@@

We identified three syncytial neurons with multiple nuclei in the aboral organ of five-day-old *M. leidyi* cydippid larvae. Burkhurdt et al. previously reported that in one-day-old larvae, the sub-epithelial nerve net (SNN) contains neurons that form syncytia [@Burkhardt2023]. However, the newly discovered syncytial neurons embedded within the aboral organ cell cluster did not exhibit the characteristic "pearl necklace-like structure" of the SNN, which consists of microtubules connecting neurons. Instead, their morphology was more reminiscent of fiber cells previously described in Placozoa [@Mayorova2021], with cell bodies extending through intercellular spaces. These morphological characteristics suggest that the newly identified neurons represent a second distinct type of syncytial neuron in ctenophores.

@@

In the this study, we found that three neurons are closely positioned within the confined space of the aboral organ, potentially enabling a single neuron to control distant ciliary movements in a synchronized manner. Moreover, the presence of neurons projecting to the same ciliated cells in distinct patterns suggests that these neurons may serve opposing functions—one inhibiting ciliary activity while the other promotes it, or one modulating ciliary beat frequency. While our synaptic ultrastructure analysis did not allow us to distinguish excitatory from inhibitory neurons, future electrophysiological investigations will be essential to elucidate their functional properties.

@@

Previous electron microscopy studies have suggested the presence of various sensory cells, including photoreceptor and mechanoreceptor cells, within the aboral organ of ctenophores [@HORRIDGE1965; @vinnikov1974e; @Krisch1973; Aronova1974]. By comparing our dataset with previously reported ultrastructural data, we successfully identified several sensory cells within the aboral organ. However, notably, we did not find any clear presynaptic structures between these sensory cells and either neural or other cellular targets. In prior studies, immunostaining for specific neuropeptides in ctenophores has revealed positive signals in distinct cell populations within the aboral organ [@Sachkova2021; @Hayakawa2022]. These findings suggest that many sensory cells in the aboral organ may primarily rely on neuropeptide-mediated paracrine signaling rather than direct synaptic connections, supporting the chemical brain hypothesis [@Jékely2021]. While paracrine transmission is generally slower than synaptic transmission, it allows for a greater diversity of information processing.

@@

Surprisingly, our study identified bridge cells as the primary cell type forming synaptic inputs onto neurons in the aboral organ. Bridge cells were first described by Tamm et al. based on their morphological characteristics, with cellular projections extending to the basal region of balancer cilia [@Tamm_2002]. Our findings suggest that bridge cells are electrically coupled to balancer cilia and provide rapid feedback to neurons via synaptic transmission. Future studies analyzing differential responses to various sensory stimuli will not only elucidate the functional roles of bridge cells but also provide insights into the evolutionary relationship between transmission speed and synapse development.

@@

The neural cells identified in the aboral organ exhibit a divergent feedforward connectivity pattern, where a single neuron forms presynaptic connections with multiple downstream targets. Such a circuit design minimizes unwanted signal interference, reduces processing delays, and enables relatively fast response execution [@Luo2021]. Moreover, the presence of multiple divergent feedforward circuits suggests that slight variations in the synchronization and phase delay of ciliary beating (within the range of a few hundred milliseconds) may have evolved as a mechanism for generating behavioral diversity.
